Supplement A: Community Clustering

### 2024-02-14 R/community-clustering.qmd
add sensitivity analysis

@@

library(arrow)
library(cppRouting)
library(ggfx)
library(ggnewscale)
library(ggspatial)
library(gt)
library(here)
library(igraph)
library(magick)
library(patchwork)
library(sf)
library(terra)
library(tidyterra)
library(tidyverse)

@@

region_labels <- as_labeller(c(
  "cmv" = "Central Mesa Verde",
  "nrg" = "Northern Rio Grande",
  "cib" = "Cibola"
))

@@

# trim all white space around image, but then add a few white pixels back (dx, dy)
prepare_image <- function(x, dx = 2, dy = 30, color = "white") {

  img <- magick::image_read(path = x)

  img <- magick::image_trim(img)

  info <- magick::image_info(img)

  new_width <- info$width + dx
  new_height <- info$height + dy

  img <- magick::image_extent(
    img, 
    geometry_area(new_width, new_height), 
    color = color
  )

  magick::image_write(img, path = x)

@@

@@

build_region <- function(gpkg, region){

@@

  if ("communities" %in% st_layers(gpkg)$name){

    community_query <- paste0("SELECT * FROM communities WHERE region = '", region, "'")
    t <- read_sf(gpkg, query = community_query) |> st_filter(r)

    t <- NULL

  list(
    name = region,
    preferred_projection = ifelse(region == "nrg", 26913, 26912),
    region = r,
    sites = s,
    communities = t
  )

add_elevation <- function(x){

  dem_fn <- paste0("dem-", x$name, ".tif")
  x$dem <- here("data", dem_fn) |> rast()

  return(x)

# getters
get_farms <- function(x){

  x$sites |> 
    st_drop_geometry() |> 
    filter(center == 0)

get_centers <- function(x){

  x$sites |> 
    st_drop_geometry() |> 
    filter(center == 1)

@@

cmv <- build_region(gpkg, "cmv") |> add_elevation()
nrg <- build_region(gpkg, "nrg") |> add_elevation()
cib <- build_region(gpkg, "cib") |> add_elevation()

@@

e <- function(x){

  tibble(
    region = x$name,
    min =  terra::global(x$dem, min, na.rm = TRUE)$min,
    max =  terra::global(x$dem, max, na.rm = TRUE)$max,
    mean = terra::global(x$dem, mean, na.rm = TRUE)$mean,
    sd =   terra::global(x$dem, sd, na.rm = TRUE)$sd
  )

bind_rows(
  e(cmv),
  e(nrg),
  e(cib)
) |> 
  gt() |> 
  tab_header("Elevation") |> 
  fmt_number(-region, decimals = 1) |> 
  opt_align_table_header("left")

@@

plot_region <- function(x){

@@

  h <- hell(x$dem)

>>>>

```{r}
#| echo: false

# allow for timing code chunks
# see https://bookdown.org/yihui/rmarkdown-cookbook/time-chunk.html
knitr::knit_hooks$set(time_it = local({

  function(before, options) {

    if(before) {

      assign('.start_time', Sys.time(), envir = .GlobalEnv)

      proc_time <- difftime(Sys.time(), .start_time, units = "secs")
      proc_time <- round(as.numeric(proc_time), 2)
      proc_time <- lubridate::duration(proc_time, "seconds")

      style <- "'text-align: right; font-size: 70%; color: rgba(55,58,60,.75)'"

      txt <- paste0("Code execution time: ", proc_time)

      paste0("<p style=", style, ">", txt, "</p>")

}))

@@

library(cppRouting)
library(furrr)
library(ggfx)
library(ggh4x)
library(ggnewscale)
library(ggspatial)
library(gt)
library(here)
library(igraph)
library(magick)
library(patchwork)
library(sf)
library(terra)
library(tidyterra)
library(tidyverse)

# cppRouting uses as.character() on node IDs, which is fine,
# but as.character(1000000) leads to "1e-6" instead of "1000000", and
# this can lead to issues when matching IDs
# to fix this, just remove the use of scientific notation
options(scipen = 9999)

# suppress progress bar and other terra output
terra::terraOptions(progress = 0, print = FALSE)

# for the sensitivity analysis, we run the entire workflow for each region in 
# its own R session, hopefully to speed things up
plan(multisession, workers = 3)

@@

region_labels <- c(
  "cmv" = "Central Mesa Verde",
  "nrg" = "Northern Rio Grande",
  "cib" = "Cibola"
)

pretty_labels <- as_labeller(c(
  "n_communities" = "No. of Communities",
  "commute" = "Mean Commute to Nearest Center [mins]",
  "room_density" = "Room Density [N/sq.km]",
  "farm_area" = "Mean Area for Farms [sq.km/N]",
  "center_area" = "Mean Area for Centers [sq.km/N]",
  "farm_ratio" = "Ratio of Farms [N] to Centers [N]",
  "area" = "Total Area [sq.km]",
  "djoin" = "D-join [mins]",
  "dmax" = "D-max [mins]",
  "p" = "p [proportion in D-join]",
  "agg_factor" = "Grid Resolution [m]"
))

squished_labels <- as_labeller(c(
  "n_communities" = "No. of\nCommunities",
  "commute" = "Mean Commute\nto Nearest Center\n[mins]",
  "room_density" = "Room Density\n[N/sq.km]",
  "farm_area" = "Mean Area\nfor Farms [sq.km/N]",
  "center_area" = "Mean Area\nfor Centers [sq.km/N]",
  "farm_ratio" = "Ratio of\nFarms [N] to Centers [N]",
  "area" = "Total Area\n[sq.km]",
  "djoin" = "D-join\n[mins]",
  "dmax" = "D-max\n[mins]",
  "p" = "p\n[prop. in D-join]",
  "agg_factor" = "Grid Resolution\n[m]"
))

@@

# trim all white space around image, 
# but then add a few white pixels back (dx, dy)
# all done with the magic of {magick}
prepare_image <- function(x, dx = 2, dy = 30, color = "white") {

  img <- image_read(path = x)

  img <- image_trim(img)

  info <- image_info(img)

  new_width <- info[["width"]] + dx
  new_height <- info[["height"]] + dy

  img <- image_extent(
    img, 
    geometry_area(new_width, new_height), 
    color = color
  )

  image_write(img, path = x)

@@

The necessary data for the analysis include a polygon defining the region of
interest, the point locations of farms and centers, and a digital elevation
model (DEM).

@@

build_region <- function(region){

@@

  dem <- here("data", paste0("dem-", region, ".tif")) |> rast()

  region_list <- list(
    region = r,
    sites = s,
    dem = dem
  )

  # save metadata
  prj <- ifelse(region == "nrg", 26913, 26912)

  class(region_list) <- "region"

  attr(region_list, "name") <- region
  attr(region_list, "preferred_projection") <- prj
  attr(region_list, "grid_crs") <- st_crs(crs(dem))$epsg

  region_list

@@

```{r}
#| eval: false
#| include: false

# this is just for my own use while running this code interactively
print.region <- function(x, ...){

  attrs <- attributes(x)

  n_farms <- x[["sites"]] |> filter(center == 0) |> nrow()
  n_centers <- x[["sites"]] |> filter(center == 1) |> nrow()

  list_stuff <- map(attrs[['names']], \(.y){

    has_class <-  class(x[[.y]])[1]

    paste0(.y, " (", has_class, ")")

  }) |> unlist()

  cli::cli_h2("Study Area Metadata")
  cli::cli_text("Region: {attrs[['name']]}")
  cli::cli_text("No. of farms: {n_farms}")
  cli::cli_text("No. of centers: {n_centers}")
  cli::cli_h3("Spatial Data")
  cli::cli_text("Grid CRS: {attrs[['grid_crs']]}")
  cli::cli_text("Preferred Projection: {attrs[['preferred_projection']]}")
  cli::cli_text("Aggregation Factor: {attrs[['agg_fact']]}")
  cli::cli_text("Max Slope: {attrs[['max_slope']]}")
  cli::cli_h3("Clustering Parameters")
  cli::cli_text("D-max: {attrs[['dmax']]} s")
  cli::cli_text("No. of outliers: {attrs[['n_outliers']]}")
  cli::cli_text("D-join: {attrs[['djoin']]} s")
  cli::cli_text("P: {attrs[['proportion']]}")
  cli::cli_text("Min. no. of farms: {attrs[['min_size']]}")
  cli::cli_h3("Community Data")
  cli::cli_text("No. of communities: {attrs[['n_communities']]}")
  cli::cli_text("No. dropped: {attrs[['n_dropped']]}")
  cli::cli_text("Concavity: {attrs[['concavity']]}")
  cli::cli_text("Dilation: {attrs[['dilate']]} m (out, in)")
  cli::cli_text("")
  cli::cli_text("List includes: {list_stuff}")

```{r}
#| time_it: true

cmv <- build_region("cmv")
nrg <- build_region("nrg")
cib <- build_region("cib")

@@

list(cmv, nrg, cib) |> 
  map(\(x){

    tibble(
      region = attr(x, "name"),
      min =  global(x[["dem"]], min, na.rm = TRUE)$min,
      max =  global(x[["dem"]], max, na.rm = TRUE)$max,
      mean = global(x[["dem"]], mean, na.rm = TRUE)$mean,
      sd =   global(x[["dem"]], sd, na.rm = TRUE)$sd
    )

  }) |> 
  bind_rows() |> 
  gt() |> 
  tab_header("Elevation") |> 
  fmt_number(-region, decimals = 1) |> 
  opt_align_table_header("left")

@@

plot.region <- function(x, ...){

  slope <- terrain(x[["dem"]], "slope", unit = "radians")
  aspect <- terrain(x[["dem"]], "aspect", unit = "radians")

  hill <- shade(slope, aspect, 30, 45)
  hill <- setValues(hill, scales::rescale(values(hill), to = c(1, 1000)))
  hill <- round(hill)

  ggplot() +
    geom_spatraster(
      data = hill, 
      fill = hcl.colors(1000, "Grays")[values(hill)],
      maxcell = Inf
    ) +
    geom_spatraster(data = x[["dem"]], alpha = 0.5) +
    scale_fill_hypso_c(
      name = "Elevation (m)",
      palette = "utah_1",
      limits = c(1350, 3850),
      alpha = 0.5,
      na.value = "transparent"
    ) +
    with_outer_glow(
      geom_sf(
        data = x[["sites"]] |> filter(center == 0),
        shape = 16,
        color = "gray25",
        alpha = 0.5,
        size = 0.6
      ),
      colour = "white",
      sigma = 0.5,
      expand = 1
    ) +
    with_outer_glow(
      geom_sf(
        data = x[["sites"]] |> filter(center == 1),
        shape = 17,
        color = "#CD3C00",
        alpha = 0.8,
        size = 1.1
      ),
      colour = "white",
      sigma = 0.5,
      expand = 1
    ) +
    annotation_scale(
      aes(location = "bl", line_col = "white", text_col = "white"),
      height = unit(0.18, "cm"),
      line_width = 0.5
    ) +
    coord_sf(expand = FALSE) +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.title = element_blank(),
      legend.position = "none",
      plot.margin = margin(l = 2)
    )

gg <- (plot(cmv) + ggtitle("Central Mesa Verde")) +
  (plot(nrg) + ggtitle("Northern Rio Grande")) + 
  (plot(cib) + ggtitle("Cibola"))

remove(plot.region, gg, fn)

It's important to note that throughout we track movement between grid cells, not
site points. So, all site points that fall into the same grid cell are assumed
to be the same location, and if any of those sites are centers, then the whole
place is considered a center.

Before converting the grid to a graph, we aggregate grid cells. Aggregating
reduces the resolution of the grid and by extension the number of grid cells the
raster contains. Since the grid cells are nodes in the grid graph, aggregating
simplifies the graph as well. This speeds up the path analysis considerably, but
it does come at the cost of precision in terms of path analysis across the
landscape, but that is not as bad as it sounds. The idea that we could build a
model that is perfectly isomorphic with the actual or true path taken by people
a thousand years ago is wildly implausible. The idea of the 'true path' is also
problematic as we're not evaluating a single commute, but rather the path people
should take more often than not. So, the real question is what level of
precision is sufficient for evaluating that. And the answer is, of course, it
just depends.

Are you walking blindfolded along the edge of a deep, precipitous cliff? Then
centimeters are probably important to you - really, really important to you. But
if you're just trying to get a reasonable estimate of the agricultural catchment
of an entirely pedestrian population based on the distribution of farms around
urban centers, probably the difference between \~30 m and \~80 m grid cells
isn't that big of a deal. If that doesn't convince you, though, we do some
limited sensitivity analysis down below.

aggregate_grid <- function(x, .factor){

  if (.factor > 1){

    x[["dem"]] <- terra::aggregate(
      x[["dem"]], 
      fact = .factor,
      fun = "mean",
      na.rm = TRUE
    )

  attr(x, "agg_fact") <- .factor

  x

group_sites_by_cell <- function(x){

  # get raster cell IDs for site points
  x[["sites"]] <- x[["sites"]] |> 
    mutate(cell = cells(x[["dem"]], vect(x[["sites"]]))[,2])

  # group relevant site attribute data by grid cell
  x[["cell_table"]] <- x[["sites"]] |> 
    st_drop_geometry() |> 
    select(cell, center, n_room) |> 
    group_by(cell) |> 
    summarize(
      center = ifelse(sum(center) > 0, 1, 0),
      n_room = sum(n_room)
    )

  x

add_graph <- function(x, max_slope = NULL){

  # keep track of metadata
  attr(x, "max_slope") <- max_slope

  # cells(<SpatRaster>) returns cell IDs 
  # for all cells with non-NA values
  non_NA_cells <- cells(x[["dem"]])

  # cells(<SpatRaster>, <numeric>) returns cell IDs 
  # for all cells with values matching <numeric>
  NA_cells <- cells(x[["dem"]], NA_real_)[[1]]

  # get adjacency matrix 
  graph <- adjacent(
    x[["dem"]],
    cells = non_NA_cells,
    directions = 8,
    pairs = TRUE
  ) |> as.data.frame()

  # remove rows with to-cells that have NA values
  graph <- graph |> filter(!to %in% NA_cells)

  # rise = difference between adjacent cells
  # run = distance between adjacent cells
  # slope = rise/run
  # convert slope 
  #   to radians with rads = atan(rise/run)
  #   to degrees with degs = rads * 180/pi 
  graph <- graph |> 
    mutate(
      from_elevation = terra::extract(x[["dem"]], from)[,1],
      to_elevation = terra::extract(x[["dem"]], to)[,1],
      rise = from_elevation - to_elevation,
      run = distance(
        xyFromCell(x[["dem"]], from),
        xyFromCell(x[["dem"]], to),
        lonlat = is.lonlat(x[["dem"]]),
        pairwise = TRUE
      ),
      slope = atan(rise/run) * (180/pi)
    )

  # what is a realistic slope people would try to hike?
  # remove all edges with slopes greater than max_slope
  if (!is.null(max_slope)) { graph <- graph |> filter(slope <= max_slope) }

  # now that we have slope, the strategy is this:
  # 1) calculate the hiking speed on each slope estimate
  # 2) calculate the length of each slope (the length of the hypotenuse)
  # 3) divide length by speed to get travel time (this is the edge weight)

  # Campbell's hiking function, see Campbell et al 2019
  # note: campbell function wants degrees!
  campbell <- function(x) {

    # values provided in Supplement Table S3
    # using the 5th percentile of the Lorentz curve as recommended by Campbell
    a <- -1.527
    b <- 14.041
    c <- 36.813
    d <- 0.320
    e <- -0.00273

    # lorentz distribution
    lorentz <- (1 / ((pi * b) * (1 + ((x - a) / b)^2)))

    # modified lorentz
    (c * lorentz) + d + (e * x)

  # calculate 
  #   speed using Campbell's hiking function
  #   length of hypotenuse
  #   travel time between adjacent cells
  graph <- graph |> 
    mutate(
      speed = campbell(slope),
      cost = sqrt(run^2L + rise^2L)/speed
    )

  # get coordinates table with cell IDs for safe joins
  coords <- data.frame(ID = non_NA_cells) |> 
    bind_cols(xyFromCell(x[["dem"]], non_NA_cells)) |> 
    rename_with(toupper)

  x[["graph"]] <- makegraph(
    graph |> select(from, to, cost),
    directed = TRUE,
    coords = coords
  )

  x

```{r}
#| time_it: true

cmv <- cmv |> 
  aggregate_grid(.factor = 3) |> 
  group_sites_by_cell() |> 
  add_graph(max_slope = 45)

nrg <- nrg |> 
  aggregate_grid(.factor = 3) |> 
  group_sites_by_cell() |> 
  add_graph(max_slope = 45)

cib <- cib |> 
  aggregate_grid(.factor = 3) |> 
  group_sites_by_cell() |> 
  add_graph(max_slope = 45)

To calculate travel time between every pair of site points, we use a
bi-directional version of Dijkstra's algorithm as implemented in the R package
`{cppRouting}`. Note that we also average the from- and to- travel times, so the
results are isotropic (meaning direction is irrelevant).

  cell_ids <- x[["cell_table"]][["cell"]]

  cells_in_graph <- x[["graph"]][["dict"]][["ref"]]

  missing_cells <- setdiff(cell_ids, cells_in_graph)

  if (length(missing_cells) > 0){

    cli::cli_abort(c(
      "Some grid cells have sites but are not in the graph!",
      "i" = "Cells: {missing_cells}"
    ))

  dm <- get_distance_matrix(
    x[["graph"]],
    from = cell_ids,  
    to = cell_ids
  )

  # make symmetric by averaging off-diagonals
  lt <- lower.tri(dm)
  ut <- upper.tri(dm)

  dm[lt] <- rowMeans(cbind(dm[lt], t(dm)[lt]), na.rm = TRUE)
  dm[ut] <- t(dm)[ut]

  # add to list and return
  x[["dm"]] <- dm

  x

```{r}
#| time_it: true

randomize_commutes <- function(x, n = 1000){

  # subset dm (distance matrix) to get dm[farms, centers]
  farms <- x[["cell_table"]] |> filter(center == 0)
  centers <- x[["cell_table"]] |> filter(center == 1)

  i <- which(rownames(x[["dm"]]) %in% farms[["cell"]])
  j <- which(colnames(x[["dm"]]) %in% centers[["cell"]])

  dm <- x[["dm"]][i, j]

  # get observed density
  observed <- apply(dm, 1, min)
  observed <- density(
    observed,
    kernel = "gaussian", 
    n = 512, 
    from = 0, 
    to = 7200
  )

  # build raster to randomly sample cells
  r <- setValues(x[["dem"]], NA)

  # don't sample past furthest observed distance from nearest center
  dmax <- st_distance(
    x[["sites"]] |> filter(center == 0), 
    x[["sites"]] |> filter(center == 1)
  )

  dmax <- units::drop_units(dmax)
  dmax <- apply(dmax, 1, min)
  dmax <- max(dmax)

  sample_area <- x[["sites"]] |> 
    filter(center == 1) |> 
    st_buffer(dist = dmax) |> 
    st_union() |> 
    vect()

  i <- cells(x[["dem"]], sample_area)[, "cell"]

  # only use those that have nodes in the graph (excludes big slopes)
  i <- i[i %in% x[["graph"]][["dict"]][["ref"]]]

  # don't sample cells with centers in them
  j <- unique(centers[["cell"]])

  random_points <- spatSample(
    r, 
    size = n,
    method = "regular",
    cells = TRUE,
    na.rm = TRUE
  )

  random_points <- unique(random_points[["cell"]])

  dm <- get_distance_matrix(
    x[["graph"]],
    from = random_points,  
    to = colnames(dm)
  )

  # make symmetric by averaging off-diagonals
  lt <- lower.tri(dm)
  ut <- upper.tri(dm)

  dm[lt] <- rowMeans(cbind(dm[lt], t(dm)[lt]), na.rm = TRUE)
  dm[ut] <- t(dm)[ut]

  expected <- apply(dm, 1, min)
  expected <- density(
    expected,
    kernel = "gaussian", 
    n = 512, 
    from = 0, 
    to = 7200
  )

  tibble::tibble(
    region = attr(x, "name"),
    distance = observed[["x"]],
    observed = observed[["y"]],
    expected = expected[["y"]]
  )

commutes <- list(cmv, nrg, cib) |> 
  map(randomize_commutes) |> 
  bind_rows()

  # subset dm (distance matrix) to get dm[farms, centers]
  farms <- x[["cell_table"]] |> filter(center == 0)
  centers <- x[["cell_table"]] |> filter(center == 1)

  i <- which(rownames(x[["dm"]]) %in% farms[["cell"]])
  j <- which(colnames(x[["dm"]]) %in% centers[["cell"]])

  dm <- x[["dm"]][i, j]

  quant <- quantile(observed, p = p)
  quant <- unname(quant)

  tibble::tibble(
    region = attr(x, "name"),
    distance = quant
  )

q95 <- list(cmv, nrg, cib) |> 
  map(\(x){ q(x, 0.95) }) |> 
  bind_rows() |> 
  mutate(
    y = c(2.4, 1.8, 1.2),
    label = case_when(
      region == "cib" ~ "Cibola",
      region == "cmv" ~ "Central Mesa Verde",
      region == "nrg" ~ "Northern Rio Grande",
      TRUE ~ NA
    )
  )

gg <- ggplot() +
  geom_hline(
    yintercept = 0,
    color = "gray75",
    linewidth = 0.2
  ) +
  geom_line(
    data = commutes, 
    aes(distance/60, log(observed/expected), color = region), 
    linewidth = 0.9,
    lineend = "round"
  ) +
  geom_point(
    data = q95,
    aes(distance/60, y, color = region),
    size = 2
  ) +
  geom_text(
    data = q95,
    aes(distance/60, y, color = region, label = label),
    vjust = 0.5,
    hjust = 0,
    nudge_x = 2,
    size = 11.5/.pt
  ) +
  scale_color_manual(name = NULL, values = region_colors) +
  annotate(
    "point",
    x = min(q95[["distance"]])/60,
    y = 3.0,
    color = "gray25",
    size = 2
  ) +
  annotate(
    "text",
    label = "Observed 95% Quantile",
    x = min(q95[["distance"]])/60 + 2,
    y = 3.0,
    color = "gray25",
    vjust = 0.5,
    hjust = 0,
    size = 11.5/.pt
  ) +
  labs(
    x = "Commute to Nearest Center [mins]",
    y = "Density\nlog(Observed) - log(Random)"
  ) +
  scale_x_continuous(
    breaks = seq(0, 120, by = 30), 
    expand = expansion(0.03)
  ) +
  coord_cartesian(xlim = c(0, 120)) +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_line(linewidth = 0.2, color = "gray85")
  )

ggsave(
  plot = gg,
  filename = fn,
  width = 5.1,
  height = 3.3,
  dpi = 600
)

remove(randomize_commutes, commutes, q, q95, gg)

Difference between log probability density of observed commute times from farms
to nearest centers and log probability density of commute times from random
locations to nearest centers.
:::

1.  Identify community centers
2.  Remove distant farms
3.  Join centers to their overlapping neighbors
4.  Join farms to their nearest center
5.  Draw smallest concave hull encompassing all farms, centers, and paths

Step 1 is already done and stored in the data. Step 2 can technically be done
after 2 and 3, but it makes sense to do it first, since it means those sites can
be ignored afterwards.

### Filter out distant farms

remove_distant_farms <- function(x, dmax){

  # subset dm (distance matrix) to get dm[farms, centers]
  farms <- x[["cell_table"]] |> filter(center == 0)
  centers <- x[["cell_table"]] |> filter(center == 1)

  i <- which(rownames(x[["dm"]]) %in% farms[["cell"]])
  j <- which(colnames(x[["dm"]]) %in% centers[["cell"]])

  dm <- x[["dm"]][i, j]

  # if the distance from a farm to its nearest center
  # is greater than dmax, we remove it from the cell table
  min_distance <- apply(dm, 1, min)

  dm <- dm[min_distance <= dmax, ]

  x[["cell_table"]] <- x[["cell_table"]] |> 
    filter(cell %in% c(rownames(dm), colnames(dm)))

  # collect metadata
  attr(x, "dmax") <- dmax
  attr(x, "n_outliers") <- nrow(farms) - nrow(dm)

  x

```{r}
#| time_it: true

cmv <- cmv |> remove_distant_farms(dmax = one_hour)
nrg <- nrg |> remove_distant_farms(dmax = one_hour)
cib <- cib |> remove_distant_farms(dmax = one_hour)

### Join neighboring centers

> For any two centers $C_1$ and $C_2$ with catchment room counts $N_1 < N_2$, if
> $pN_1$ of $C_1$'s rooms are within a distance $D$ of $C_2$, then $C_1$ is part
> of the same community as $C_2$.

Here the population, $N$, of focal center $c$ is defined as

$$N_{c} = R_{c} + \sum_{k=1}^{S} R_{k} \cdot I(t_{kc} \leq D_{c})$$

for all sites $S$ (this includes both farms and centers), with $R_{c}$ being the
room count at $c$ and $R_{k}$ the room count at site $k$, $t_{kc}$ the travel
time from $k$ to $c$, $D_{c}$ the threshold travel time to $c$ (referred to as
*D-join* in the paper), and $I$ the indicator function that is 1 if $t \leq D$
and 0 otherwise.

  # subset dm (distance matrix) to get dm[farms and centers, centers]
  center_cells <- x[["cell_table"]] |> filter(center == 1) |> pull(cell)

  j <- which(colnames(x[["dm"]]) %in% center_cells)

  dm <- x[["dm"]][, j]

  # find farms and centers that are within distance djoin of other centers
  djoins <- apply(dm, 1, \(.x){ which(.x <= djoin) })

  # remove farms that are isolated (ie, greater than djoin from a center)
  # note: this should just be farms because all centers are less than
  #   djoin from themselves
  djoins <- djoins[lengths(djoins) > 0]

  # collect list into a long-form table
  djoins <- tibble(
    center = colnames(dm)[unlist(djoins)],
    farm_or_center = rep(names(djoins), lengths(djoins))
  ) |> mutate(across(everything(), as.numeric))

  # now we combine the room counts with the djoins table
  djoins <- djoins |> 
    left_join(
      x[["cell_table"]] |> select(-center), 
      by = join_by(farm_or_center == cell)
    ) |> 
    rename("population" = n_room)

  # calculate the total population of the catchment around each center
  # note that this involves a lot of double-counting, farms and centers
  # counting towards the population size of multiple centers, but this is OK
  # as we are interested in using the shared population as a measure of 
  # interaction between centers, not as an estimate of a community's population
  gen_pop <- djoins |> 
    group_by(center) |> 
    summarize(population = sum(population))

  # for each center, what other centers have larger catchment populations?
  # using outer() is extremely fast
  # in this case, it creates an asymmetric logical matrix
  R <- outer(gen_pop[["population"]], gen_pop[["population"]], FUN = '<=')
  diag(R) <- FALSE

  colnames(R) <- gen_pop[["center"]]
  rownames(R) <- gen_pop[["center"]]

  # now, we turn that matrix into a big list 
  # each element is a vector with the cell IDs of the larger centers
  R <- apply(R, 1, which)
  R <- map(R, \(x){ as.numeric(names(x)) })

  # for each center, we look at all the centers with bigger catchment
  # populations than it, and ask: is the shared population size some fraction
  # of the total population of the catchment?
  neighbors_matrix <- purrr::imap(R, \(x, idx){

    N <- gen_pop |> 
      filter(center == as.numeric(idx)) |> 
      pull(population)

    focal_neighbors <- djoins |> 
      filter(center == as.numeric(idx)) |> 
      pull(farm_or_center)

    cntrs <- djoins |> 
      filter( # this filter gives the intersection of farms/cells 
        center %in% x,
        farm_or_center %in% c(idx, focal_neighbors)
      ) |> 
      group_by(center) |> 
      summarize(proportion = sum(population)/N) |> 
      filter(proportion >= p) |> 
      pull(center)

    # return a logical vector the length of R
    names(R) %in% cntrs

  # there's a recursive dimension to this:
  # if a center is joined with a center that's joined with a center, etc.
  # i don't know how to do that in R
  # so, igraph it is...
  g <- igraph::graph.adjacency(neighbors_matrix)
  membership <- igraph::components(g)[["membership"]]

  # this gives us the community that every center is part of
  x[["center_communities"]] <- tibble(
    center = names(membership) |> as.numeric(),
    community = unname(membership)
  )

  # collect metadata
  attr(x, "djoin") <- djoin
  attr(x, "n_communities") <- length(unique(unname(membership)))
  attr(x, "proportion") <- p

  x

```{r}
#| time_it: true

### Collect members

Here we just associate each farm with its nearest community center. So, we have
full community membership defined. Then we drop communities that have three or
less sites associated with them. This is only done because our main goal with
this analysis is to estimate the extent of farm land associated with these
dispersed farming communities.

  # subset dm (distance matrix) to get dm[farms, centers]
  farms <- x[["cell_table"]] |> filter(center == 0)
  centers <- x[["cell_table"]] |> filter(center == 1)

  i <- which(rownames(x[["dm"]]) %in% farms[["cell"]])
  j <- which(colnames(x[["dm"]]) %in% centers[["cell"]])

  dm <- x[["dm"]][i, j]

  # get nearest center to each farm
  k <- apply(dm, 1, which.min)
  nearest_centers <- tibble(
    farm = as.numeric(names(k)),
    nearest_center = as.numeric(colnames(dm)[k])
  )

  # collect the farms and the centers they are closest to, 
  # and by extension the communities they are part of
  farms <- x[["cell_table"]] |> 
    filter(center == 0) |> 
    left_join(
      nearest_centers, 
      by = join_by(cell == farm)
    ) |> 
    left_join(
      x[["center_communities"]], 
      by = join_by(nearest_center == center)
    ) |> 
    select(-nearest_center)

  # collect the centers and their community affiliations
  centers <- x[["cell_table"]] |> 
    filter(center == 1) |> 
    left_join(
      x[["center_communities"]], 
      by = join_by(cell == center)
    )

  # combine and save
  x[["cell_table"]] <- bind_rows(farms, centers)

  x

  final_list <- x[["cell_table"]] |> 
    filter(center == 0) |> 
    group_by(community) |> 
    summarize(n = n()) |> 
    filter(n > threshold) |> 
    pull(community) |> 
    unique()

  x[["cell_table"]] <- x[["cell_table"]] |> filter(community %in% final_list)

  # update metadata
  attr(x, "min_size") <- threshold
  attr(x, "n_dropped") <- attr(x, "n_communities") - length(final_list)

  x

```{r}
#| time_it: true

cmv <- cmv |> 
  collect_members() |> 
  drop_small_communities(threshold = 3)

nrg <- nrg |> 
  collect_members() |> 
  drop_small_communities(threshold = 3)

cib <- cib |> 
  collect_members() |> 
  drop_small_communities(threshold = 3)

Now that we have community centers joined to their overlapping neighbors, and
farms joined to their nearest centers, we can define boundaries for the
resulting communities using a concave hull. We choose a concave rather than
convex hull because the latter tends to artificially inflate the size of the
community, often to a significant degree. On the other hand, the concave hull
can generate extremely unrealistic shapes for communities, even when moving the
concavity ratio closer to the convex hull.

Our somewhat brute force strategy for navigating between the Scylla and
Charybdis of being too big or too strange is to first compute the least-cost
path between every site on the concave hull of the community's site points. In
this case the concave hull gives us all the points on the convex hull along with
some on the interior of the convex hull. The points that are excluded should
have shortest paths to the outer points that at some point intersect with the
shortest paths that are actually calculated, so we get a reasonable
approximation and save some time doing it. We then add the vertices that define
the computed paths as virtual points to the community and generate the concave
hull for the expanded set of community points. The consequence is that an
individual setting off from one farm in the community to any other site in the
community (farm or center) should never have to leave the community. This, in
effect, makes the community polygons concave with respect to time.

The result can still be a little noisy, so we simplify polygons - including, in
some case, filling holes - by dilating them, that is, adding then removing a
small buffer.

Note that we also do some graph pruning in each iteration, basically cropping
the graph to an expanded area around the points assigned to a specific
community. This prevents the search algorithm from searching out away from the
other sites in the community.

  members <- x[["cell_table"]]

  sites <- x[["sites"]] |> 
    select(cell) |> 
    st_transform(attr(x, "grid_crs"))

  community_ids <- unique(members[["community"]])

    m <- members |> filter(community == i)

    # get cells along the edge of point clusters using concave hull
    # so we get points that are on the convex hull and in
    # the interior of the cluster, too
    pts <- xyFromCell(x[["dem"]], m[["cell"]]) |> 
      st_multipoint() |> 
      st_sfc(crs = attr(x, "grid_crs")) |> 
      st_sf(geometry = _)

    cells <- pts |> 
      st_cast("POINT") |> 
      mutate(cell = m[["cell"]]) |> 
      st_filter(st_concave_hull(pts, 0.2), .predicate = st_touches) |> 
      pull(cell)

    # prune the graph so search doesn't go too far away from the convex hull,
    # meaning away from all the other points in the community
    # we use a somewhat arbitrary 1.5 km distance for this, so we can reasonably 
    # assume we are not chopping off the shortest path, while at the same time
    # cutting down the total number of paths to search by a considerable amount
    bb8 <- st_convex_hull(pts) |> 
      st_buffer(1500) |> 
      st_bbox() |> 
      st_as_sfc(crs = st_crs(pts))

    cells_in_bb8 <- cells(x[["dem"]], vect(bb8))[,2]

    graph <- x[["graph"]]

    node_ids <- graph[["dict"]] |> 
      filter(ref %in% cells_in_bb8) |> 
      pull(id)

    graph[["data"]] <- graph[["data"]] |> 
      filter(from %in% node_ids, to %in% node_ids)

    # get all short paths 
    # this returns a list with length equal to the length of cells
    # each element of the list is a vector of the grid cells that the short
    # path passes through
    paths <- get_multi_paths(graph, from = cells, to = cells)

    # unraveling this list gives us a vector of all the cells on shortest paths
    paths <- paths |> unlist(recursive = TRUE) |> as.numeric() |> unique()

    site_xy <- sites |> 
      filter(cell %in% m[["cell"]]) |> 
      st_coordinates()

    names(site_xy) <- tolower(names(site_xy))

    xy <- rbind(
      xyFromCell(x[["dem"]], paths),
      site_xy
    )

    border <- xy |> 
      st_multipoint() |> 
      st_sfc(crs = attr(x, "grid_crs")) |> 
      st_sf(geometry = _) |> 
      st_concave_hull(ratio = concavity_ratio) |> 
      mutate(community = i)

  x[["communities"]] <- communities |> 
    bind_rows() |> 
    st_transform(attr(x, "preferred_projection")) |> 
    st_buffer(dilate[1]) |> 
    st_buffer(dilate[2]) |> 
    st_transform(4326)

  # update metadata
  attr(x, "concavity") <- concavity_ratio
  attr(x, "dilate") <- dilate

  x

```{r}
#| time_it: true

cmv <- cmv |> define_boundaries(concavity_ratio = 0.1, dilate = c(130, -100))
nrg <- nrg |> define_boundaries(concavity_ratio = 0.1, dilate = c(130, -100))
cib <- cib |> define_boundaries(concavity_ratio = 0.1, dilate = c(130, -100))

visualize_communities <- function(x){

@@

  h <- hell(x[["dem"]])

  members <- x[["cell_table"]] |> 
    group_by(community) |> 
    summarize(n_room = sum(n_room))

  communities <- x[["communities"]] |> 
    left_join(members, by = "community") |> 
    mutate(
      area = st_area(geometry) |> units::set_units(km2) |> units::drop_units(),
      density = n_room/area,
      density = log(density+1),
      density = cut(density, 9, labels = FALSE)
    )
